---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(lofifonts)
```


```{r eval=FALSE, include=FALSE}
pkgdown::build_site(override = list(destination = "../coolbutuseless.github.io/package/lofifonts"))
```

# lofifonts


<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![R-CMD-check](https://github.com/coolbutuseless/lofifonts/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/lofifonts/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/lofifonts)](https://CRAN.R-project.org/package=lofifonts)
<!-- badges: end -->

`lofifonts` provides access to raw data for the included bitmap and vector fonts.

Bitmap and vector fonts allow for bespoke rendering of pixels and strokes in 
situations where fine control over rendering is required.  For example, 
individual pixels and strokes could be animated to create novel effects like
exploding or coalescing text.  With access to coordinates, variations in 3D
also become possible.

This package will render text with a specified bitmap or vector font to:

* a data.frame of coordinates (for bitmap fonts)
* a data.frame of character strokes (for vector coordinates)
* a binary matrix indicating the rendered text
* a raster image of the rendered text



### What's in the box

* Bitmap font functions
    * `bitmap_text_coords()` create a data.frame of pixel locations
    * `bitmap_text_matrix()` create a binary matrix with pixel locations set to 1
    * `bitmap_text_raster()` create a raster image of the text
* Vector font functions
    * `vector_text_coords()` create a data.frame of strokes
    * `vector_text_matrix()` create a binary matrix with pixel locations set to 1
    * `vector_text_raster()` create a raster image of the text

### Fonts


| Type   | Name            | Sizes                         | Unicode?              | # glyphs |
|--------|-----------------|-------------------------------|-----------------------|----------|
| Bitmap | Spleen          | 5x8, 6x12, 8x16, 12x24, 16x32 | Some                  | 450-1000 |
| Bitmap | Tamzen          | 5x9, 6x12, 7x13, 8x15, 8x16, 10x20 bold/regular | Some |  180    |
| Bitmap | Unifont         | 16x16                         | Yes. Plane 0, 1       | 113446   |
| Bitmap | Unscii          | 8x8, 8x16                     | Some                  | 3240     |
| Vector | gridfont        |                               | Lower case ASCII only | 68       |
| Vector | gridfont_smooth |                               | Lower case ASCII only | 68       |
| Vector | arcade          |                               | Upper case ASCII only | 66       |


Bitmap fonts:

* [Spleen v2.1.0](https://github.com/fcambus/spleen) License: BSD 2-clause. See `LICENSE-spleen.txt`
* [Tamzen v1.11.6](https://github.com/sunaku/tamzen-font) Free to distribute
* [Unifont v16.0.0.1](https://unifoundry.com/unifont/) License: SIL Open Font License (OFL) version 1.1. See `LICENSE-unifont.txt`
* [Unscii v2.0](https://github.com/viznut/unscii) Font is in the public domain.

Vector fonts

* [gridfont](https://github.com/inconvergent/gridfont) License: MIT. See `LICENSE-gridfont.txt`
* `arcade` is a vector font I created. License: SIL Open Font License (OFL) version 1.1. See `LICENSE-arcade.txt`


## Installation

You can install from [GitHub](https://github.com/coolbutuseless/lofifonts) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/lofifonts')
```

```{r echo=FALSE, eval=FALSE}
txt <- "привет"
txt <- "二項分布\xF0\x9F\x8E\xB2の英語表記は\n「Binomial distribution」である。"
```


## Included fonts

```{r}
get_lofi_names('all')
```


## Bitmap font: Rendering text

Text may be rendered with a bitmap font to 

1. A data.frame of pixel locations
2. A binary matrix of pixel locations
3. A simple `raster` object 

```{r bitmap-coords, fig.height = 3}
library(lofifonts)
bitmap_text_coords("Hello", font = 'unifont') |>
  head()
```


```{r bitmap-matrix, fig.height = 3}
bitmap_text_matrix("Hello", "unifont") 
```


```{r bitmap-raster, fig.height = 3}
bitmap_text_raster("Hello", "unifont") |>
  plot(interpolate = FALSE)
```

```{r bitmap-raster-codepoints, fig.height = 3}
# Render codepoints directly
bitmap_text_raster(intToUtf8(c(0x1f44d, 0x1f44e)), "unifont") |>
  plot(interpolate = FALSE)
```

## Bitmap font: Bespoke pixel rendering

This is an example of bespoke rendering of the pixel data for an example string.

Each pixel in the `coords` data.frame has an `x` and `y` coordinates, and to 
render this font, a small square will be drawn for each pixel.  

A gap will be left between the rectangels to highlight the pixelated origin 
of the text.

```{r bitmap-bespoke, fig.height = 4}
library(grid)
coords <- bitmap_text_coords("Hello\n#RStats", "spleen-6x12")
head(coords)

grid.newpage()
grid.rect(
  x = coords$x * 4, 
  y = coords$y * 4, 
  width  = 3, 
  height = 3, 
  default.units = 'mm',
  gp = gpar(fill = rainbow(nrow(coords)), col = NA)
)
```



## Vector font: Rendering text

Text may be rendered with a vector font to 

1. A data.frame of stroke endpoints
2. A binary matrix of pixel locations
3. A simple `raster` object 

```{r vector-coords, fig.height = 3}
library(lofifonts)
vector_text_coords("Hello") |> 
  head()
```


```{r vector-matrix, fig.height = 3}
vector_text_matrix("Hello", scale = 1) 
```


```{r vector-raster, fig.height = 3}
vector_text_raster("Hello\n#RStats!", scale = 5) |>
  plot(interpolate = FALSE)
```


## Vector font: Bespoke pixel rendering

This is an example of bespoke rendering of the strokes for an example string.

For each character (`char_idx`) there are 1-or-more strokes (`stroke`).  Each stroke
has has 2 or more rows giving the path for this particular stroke.

When plotting with ggplot, draw path for the points-within-strokes-within-characters.


```{r vector-bespoke, fig.height = 3}
library(ggplot2)

coords <- vector_text_coords("Hello")
head(coords)

ggplot(coords) +
  geom_path(aes(x, y, 
                group = interaction(char_idx, stroke_idx), 
                colour = as.factor(char_idx)), 
            linewidth = 4) +
  geom_point(aes(x, y), color = 'yellow') + 
  theme_bw() +
  theme(legend.position = 'bottom') + 
  coord_equal()
```




## Unicode support

The font with the best unicode coverage is the bitmap font `unifont`

```{r unicode, fig.height = 3}
library(lofifonts)

txt <- "二項分布\xF0\x9F\x8E\xB2の英語表記は\n「Binomial distribution」である。"

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Render quickly with built-in conversion
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
bitmap_text_raster(txt, "unifont") |>
  plot(interpolate = FALSE)
```


```{r unicode2, fig.height = 2}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Bespoke plotting of the pixel locations
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
coords <- bitmap_text_coords(txt, "unifont")

ggplot(coords) +
  geom_tile(aes(x, y, fill = as.factor(char_idx)), na.rm = TRUE) +
  theme_void(10) +
  coord_equal() + 
  theme(legend.position = 'none')
```


## Rendering with `{bittermelon}` fonts

[`bittermelon`](https://cran.r-project.org/package=bittermelon) is a package
which supports reading and manipulating many different types of bitmap font.

To use a font loaded with `bittermelon` is must first be convert to the 
`lofi` format which this package understands.

```{r}
# Load font using "bittermelon"
filename <- system.file("fonts/fixed/4x6.yaff.gz", package = "bittermelon", mustWork = TRUE)
bmfont <- bittermelon::read_yaff(filename)

# Convert to lofi format
lofi <- convert_bm_font_to_lofi(bmfont)
lofi

# Layout the text and render to a raster
bitmap_text_raster('hello', font = lofi) |>
  plot(interpolate = FALSE)
```




## Large block of text in bitmap font

```{r}
library(lorem)
set.seed(1)
oldpar <- par(mai = c(0, 0, 0, 0))

lorem::ipsum(2) |>
  strwrap(40) |>
  paste(collapse = "\n") |>
  bitmap_text_raster() |>
  plot()

par(oldpar)
```


## Large block of text in vector font

```{r fig.width = 21, fig.height = 6}
library(lorem)
set.seed(1)
oldpar <- par(mai = c(0, 0, 0, 0))

lorem::ipsum(1) |>
  strwrap(60) |>
  paste(collapse = "\n") |>
  vector_text_raster() |>
  plot(interpolate = FALSE)

par(oldpar)
```



## Bitmap Fonts: Font Sheets

A selection of the glyphs from each of the bitmap fonts

```{r fontsheets, echo=FALSE, eval=TRUE, fig.height = 4}
lo <- utf8ToInt('a')
hi <- utf8ToInt('z')
txt <- intToUtf8(seq(lo, hi))

for (font_name in get_lofi_names('bitmap')) {
  coords <- bitmap_text_coords(txt, font_name)
  
  p <- ggplot(coords) +
    geom_tile(aes(x0, y0), width=1, height = 1, na.rm = TRUE) +
    facet_wrap(~char_idx, ncol = 9)+
    theme_minimal(15) +
    coord_equal() + 
    labs(title = font_name) +
    theme(
      strip.text = element_blank(),
      axis.text  = element_blank(),
      axis.title = element_blank()
    )
  
  print(p)
}
```

## Vector fonts: Font Sheets

A selection of the glyphs from each of the vector fonts

```{r vectorsheets, echo=FALSE, eval=TRUE, fig.height = 4}
lo <- utf8ToInt('a')
hi <- utf8ToInt('z')
txt <- intToUtf8(seq(lo, hi))

for (font_name in get_lofi_names('vector')) {
  coords <- vector_text_coords(txt, font_name)
  
  p <- ggplot(coords) +
    geom_path(aes(x0, y0, group = stroke_idx)) +
    facet_wrap(~char_idx, ncol = 9)+
    theme_minimal(15) +
    coord_equal() + 
    labs(title = font_name) +
    theme(
      strip.text = element_blank(),
      axis.text  = element_blank(),
      axis.title = element_blank()
    )
  
  print(p)
}
```


```{r echo=FALSE, eval=FALSE}
set.seed(1); lorem::ipsum(2) |> strwrap(40) |> paste(collapse = "\n") -> txt
bitmap_text_raster(txt) |> bench::mark()

set.seed(1); 
txt <- lorem::ipsum(2) |> strwrap(40) 
txt <- paste(txt, collapse = "\n") -> txt

# txt <- 'hello\nworld\nthere\ngoes\nthe'
vector_text_coords(txt) |> bench::mark()

vector_text_matrix(txt, scale = 1) 
vector_text_raster(txt, scale = 3) |> plot(interpolate = FALSE)


set.seed(1); 
lorem::ipsum(2) |> 
  strwrap(40) |> 
  paste(collapse = "\n") |>
  vector_text_raster(scale = 3) |> 
  plot()
```










